---
title: "Data Types in R"
author: "Revendranath T"
date: "8/24/2021"
output: ioslides_presentation
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```




## Using R Console

- 1+1
- max(1,2,5)
- min(1,3,8,9)
- q() #exit from R & RStudio
- escape key to interrupt R
- use help() or ? to get help from R
- example help(mean or ?mean)

## Slide with R Output

```{r basics, echo = TRUE}
pi
sqrt(25)
5^2
```

## Use print
- To format any R value for printing, including structured values such as matrices and lists
```{r print, echo=false}
#print matrix

print(matrix(c(1, 2, 3, 4), 2, 2))

matrix(data=c(1, 2, 3, 4), 
        nrow=2, 
        ncol=2,
        byrow=TRUE
        )

# 3x2, 1,4,6,8,9,0
# 1,8
# 4,9
# 6,0

matrix(data=c(1,4,6,8,9,0), 
        nrow=3, 
        ncol=2,
        byrow=FALSE
        )

#print list
print(list("a", "b", "c"))

```


## Working with variables
- No need to declare variables first
- Assign variables directly through = or <-
- Global variables: <<-
- List variables: ls() or ls.str()
- Delete variables: rm()

```{r variables, echo=TRUE}
# = or <- are used for assignment

x = 3.9
y <- 4

x
y

x = 'a' #is same as
x <- 'a' #is same as
x = "a" #is same as
x <- "a" #is same as

typeof(x)  # data type
class(x)   # for data structure
mode(x)

x = c("fee", "fie", "foe", "fum") # x is a vector
print(x)

x  <-  c("fee", "fie", "foe", "fum") # x is a vector
print(x)

#global variable assignment <<-
#That forces the assignment to a global variable rather than a local variable
x <<- 3

#listing variables
ls()
ls.str()


# delete a variable
rm(x)
ls()

# delete multiple variables
rm(bm,y)
ls()
```

### Create vectors
- Use the c(...) operator to construct a vector from given values
- Use mode to know the type of vector: numeric, character, boolean

```{r vectors, echo=TRUE}
x1  <-  c(1, 1, 2, 3, 5, 8, 13, 21)
x1
mode(x1)
typeof(x1) 
class(x1) 

f1 <- c(1 * pi, 2 * pi, 3 * pi, 4 * pi)
mode(f1)
typeof(f1)

s1 <- c("Welcome", "to", "BITS", "Pilani", "!")
mode(s1)
typeof(s1)

l1 <- c(TRUE, TRUE, FALSE, TRUE)
mode(l1)
typeof(l1)

3.5

```
### Modes
- What happens when you merge vectors of two different modes?
```{r modes, echo=TRUE}

v1 <- c(1, 2, 3)
v1
mode(v1)
v3 <- c("A", "B", "C")
v3
mode(v3)
v2 <- c(v1, v3)
v2
mode(v2)
```

## Basic Statistics with R
- mean()
- median()
- var()
- sd()
- cor()
- cov()

```{r, echo=TRUE}
#assign values to variable x
x <- c(0, 1, 1, 2, 3, 5, 8, 13, 21, 34)
x
#mean
mean(x)

#median
median(x)

#variance
var(x)

#standard deviation
sd(x)

#correlation & covariance. Need 2 variables. 

# create a y variable
print("x")
print(x)
print("x+1")
print(x+1)
x + 1
y <- log(x + 1)
y
#correlation
cor(x, y)
#covariance
cov(x, y)

```

## Handling null values or NA
- exclude null values using na.rm = TRUE
- use imputation methods (discussed later)


```{r, echo=TRUE}
x <- c(0, 1, NA , NA, 1, 2, 3, NA, NA)
x
is.na(x)
#without na.rm = TRUE
mean(x) # is equivalent to mean(x, na.rm = FALSE)
sd(x)  # is equivalent to sd(x, na.rm = FALSE)

#with na.rm = TRUE
mean(x, na.rm = TRUE)
sd(x, na.rm = TRUE)

```

## Basic statistics for multiple columns
- use map_dbl() function from purrr package in tidyverse

```{r, echo=TRUE}
# use ?mtcars to verify the dataset
library(tidyverse)
data(mtcars)
str(mtcars)
head(mtcars)
#mean
map_dbl(mtcars, mean)

#median
map_dbl(mtcars, median)

#sd
map_dbl(mtcars, sd)

#var
map_dbl(mtcars, var)

#all summary statistics
summary(mtcars)

#correlation
cor(mtcars)

#covariance
cov(mtcars)
```


## Data types
* Total six data types
  * character
  * numeric (real or decimal) 
  * integer 
  * logical 
  * complex


## Data Structures
* Five prominent data structures
  * atomic vector (homogeneous)
  * matrix (homogeneous)
  * arrays (homogeneous)
  * list (heterogeneous)
  * data frame (homogeneous)

## Use the following to examine them

* class() - what kind of object is it (high-level)?
* typeof() - what is the object’s data type (low-level)?
* length() - how long is it? What about two dimensional objects?
* attributes() - does it have any metadata?
  * names
  * dimnames
  * dim
  * class

```{r}
#character
x <- "hello"
class(x)
typeof(x)
length(x)
attributes(x)
nchar(x) # kind of an attribute
```

```{r}
# integer array
x <-  1:10
x
class(x)
typeof(x)
length(x)
attributes(x)
```

```{r}
# numeric array
x <-  as.numeric(1:10)
x <- as.numeric(c(1.1, 2.3, 5.6, 12))
x
class(x)
typeof(x)
length(x)
attributes(x)
```

## Matrix operations
```{r}
matrix(c(1, 2, 3, 4), nrow = 2, ncol=2  )

x <- matrix(c(1, 2, 3, 4, 5, 6), 
            nrow = 2, 
            ncol = 3)

x
class(x)
typeof(x)
length(x)
attributes(x)
```

```{r}
rownames(x) <- c("Pilani", "Goa")
x
colnames(x) <- c("Hyderabad", "Dubai", "Mumbai")
x
attributes(x)
```

### Lists
```{r}
x <- c(1, "BITS", FALSE, 10-9i) # Is this a list or vector?

x <- list(as.integer(1), "BITS", FALSE, 10-9i) # Is this a list or vector?

##What is the difference

x
class(x)
typeof(x)

typeof(x[[1]])
typeof(x[[2]])
typeof(x[[3]])
typeof(x[[4]])

length(x)
attributes(x)
```


### Create sequences & Repetitions
- create a sequence of numbers
- Use an n:m expression to create the simple sequence n, n+1, n+2, …, m
- Use the `seq` function for sequences with an increment other than 1
- Use the rep function to create a series of repeated values

```{r sequences, echo=TRUE}
101:115

seq(from = 1, to = 22, by = 5)

rep(8, times = 5)

rep("BITS", times = 3)
```


### Dataframes
```{r}
letters[10:20]

x <- data.frame(letters1 = letters[11:20], 
                num1 = 1:10, 
                num2 = as.numeric(11:20)
                )
x

View(x)

class(x)
typeof(x$num2)
length(x)
dim(x)
attributes(x)

```

```{r}
#functions to explore dataframes
head(x,3)
tail(x,4)
dim(x)
nrow(x)
ncol(x)
str(x)
names(x)
rownames(x)
colnames(x)


```


## Comparing vectors
```{r, echo=TRUE}
a <- 3
a
pi
a == pi # Test for equality

a != pi # Test for inequality

a < pi

a > pi

a <= pi

a >= pi

```


```{r, echo=TRUE}
v <- c(3, pi, pi, 0)
v
length(v)

w <- c(pi, pi)
w
length(w)
```

```{r, echo=TRUE}
v == w # Compare two vectors with different sizes
v
v < 3.14

v <- c(3, pi, pi, 0, 3, pi)
w <- c(pi, pi) 
v != w
v < w
v <= w
v > w
v >= w
v == w
```

```{r}
1+19
```



```{r, echo=TRUE}
v <- c(3, pi, 4)
v == pi # Compare a 3-element vector against one number

v != pi

```

## Select elements in the vector
```{r, echo=TRUE}
fib <- c(0, 1, 1, 2, 3, 5, 8, 13, 21, 34)

fib

#fib[0]

fib[1]

fib[2]

fib[3]

fib[4]

fib[5]

```

```{r, echo=TRUE}
fib[1:3] # Select elements 1 through 3

fib[4:9] # Select elements 4 through 9

fib[c(1, 2, 4, 8)]

fib[-1] # Ignore first element

fib[1:3] # As before

fib[-(1:3)] # Invert sign of index to exclude instead of select

x[1:3,3]

x
```


```{r, echo=TRUE}
fib
fib < 10 # This vector is TRUE wherever fib is less than 10

fib[fib < 10] # Use that vector to select elements less than 10

fib[fib > 10]

fib %% 2 == 0 # This vector is TRUE wherever fib is even

fib[fib %% 2 == 0] # Use that vector to select the even elements
```


## Combining vector comparisons, logical operators, and vector indexing

```{r, echo=TRUE}
v <- c(3, 6, 1, 9, 11, 16, 0, 3, 1, 45, 2, 8, 9, 6, -4)
v
quantile(v)

quantile(v, 0.25)

z <- c(-4, 0, 1, 1, 2, 3, 3, 6, 6, 8, 9, 9, 11, 16, 45)

print(quantile(v, 0.1))
print(quantile(z, 0.1))


length(z)

summary(v)

boxplot(v)

v > median(v)

v[ v > median(v)]

v[ (v < quantile(v, 0.05)) | (v > quantile(v, 0.95)) ]

#Select all elements that exceed ±1 standard deviations from the mean:

v[ abs(v - mean(v)) > sd(v)]

#Select all elements that are neither NA nor NULL:

v <- c(1, 2, 3, NA, 5, 8)
is.na(v)

na.rm =TRUE

v[!is.na(v)]

v[!is.na(v) & !is.null(v)]

median(v)

quantile(v, 0.25)

desc(v)
```

## Vector Arithmetic
```{r, echo=TRUE}
v <- c(11, 12, 13, 14, 15)
w <- c(1 , 2 , 3 , 4 , 5)
v + w

v - w

v * w

v / w

w^v

```


## vector-scalar operations
```{r, echo=TRUE}
w

w + 2

w - 2

w * 2

w / 2

2^w
```

```{r, echo=TRUE}
w

#sqrt
sqrt(w)

#log
log(w)

#sin
sin(w)

#mean
mean(w)

#distance from mean
w - mean(w)

#sd
sd(w)

#z-score
(w - mean(w))/sd(w)


```

##  Choosing Between Two Alternatives: if-else

```{r, echo=TRUE}
#syntax 
#if (condition) {
  ##do this if condition is TRUE 
#} 
#else {
  ## do this if condition is FALSE
#}


#Example
```


```{r, echo=TRUE}
x <- c(4, -2, -1, 0, 1, 2)
x
sqrt(x)
#purrr package
if (x >= 0) {
  print(sqrt(x))             # do this if x >= 0
} else {
  print("negative number")   # do this otherwise
} 

x <-  8
x
# -1 & 0 # print("between -1 and zero number")
# 0 and 5 print("between zero & 5number")
# else greater than 5

ifelse(( x > -1 && x < 0), 
  #IF TRUE
       print("between -1 and zero number"),
  #IF first condition is FALSE
  ifelse((x > 0 & x < 5 ),
         #IF TRUE
         print("between zero and 5"),
         print(" greater than 5"))
         )
  
x <- 5

# if(condition 1) {
 #expressions of condition 1 is true
#} else if (condition){
  #expressions of condition 1 is false & condition 2 is true
#} else {
  #expressions of condition 1 is false and condition 2 is false
#}
  
x <-  3
x

if( x > -1 && x < 0) { 
  #IF TRUE
       print("between -1 and zero number") 
  } else if(x > 0 && x < 5)
  #IF first condition is FALSE
   {
         #IF TRUE
         print("between zero and 5")
} else {
         print("greater than 5")
}
      


```

```{r, echo=TRUE}
x <- c(-2, -1, 0, 1, 2)

if (all(x < 0)) {
  print("all are negative")
}

if (any(x < 0)) {
  print("some are negative")
}
```


##  Iterating with a Loop
```{r, echo=TRUE}
#Syntax
for (x in v) {
  # do something with x
}

y <- c(-2, -1, 0, 1, 2)
y

print(y)

for(x in y) {
  print(x) 
  print(x^2)
}

x <-  NA

x

x<- c(-2, -1, 0, 1, 2)
x
length(x)

for(x in 1:5) {
  print("length: ") 
  print(length(x))
  print(x) 
  print(x^2)
}

print(y)

x <- c(-2, 2, 0, 1, 5)
x
y <- 1:5
y

x_len <- length(x)
y_len <- length(y)

for (i in 1:x_len) {
  #print(i)
  #print(x[[i]])
  #print(y[[i]])
  if(x[[i]] == y[[i]]) {
    print(x[[i]])
    print(x[[i]]^2)
  }
}

for (i in 1:5) {
  v[[i]] <- v[[i]] ^ 2
}
print(v)

print(x)
print(y)

```


## functions
```{r, echo=TRUE}
#syntax
name <- function(param1, ..., paramN) {
          expr1
          .
          .
          .
          exprM
}

#function for coefficient of variation
x <-  1:15

cv <- function(x) {
  sd(x) / mean(x)
}

class(cv)

cv(x)
cv(1:10)     # Set x = 1:10 and evaluate sd(x)/mean(x)
```

## Three parts of R functions :
* the body(), the code inside the function.
* the formals(), the list of arguments which controls how you can call the function.
* the environment(), the “map” of the location of the function’s variables.

```{r}
body(cv)
formals(cv)
environment(cv)
```


```{r}
wt_mean <- function(x, w) {
  sum(x * w) / sum(w)
}

wt_var <- function(x, w) {
  mu <- wt_mean(x, w)
  sum(w * (x - mu) ^ 2) / sum(w)
}
wt_sd <- function(x, w) {
  sqrt(wt_var(x, w))
}

#example
wt_mean(1:6, 1:3)

#same size of x and w
wt_mean <- function(x, w) {
  if (length(x) != length(w)) {
    stop("`x` and `w` must be the same length", call. = TRUE)
  }
  sum(w * x) / sum(w)
}

```

```{r}
celsius_to_kelvin <- function(temp_C) {
  temp_K <- temp_C + 273.15
  return(temp_K)
}

# freezing point of water in Kelvin
celsius_to_kelvin(0)
```


```{r}
fahrenheit_to_kelvin <- function(temp_F) {
  #temp_C <- fahrenheit_to_celsius(temp_F)
  temp_K <- celsius_to_kelvin(temp_F)
  return(temp_K)
}

# freezing point of water in Kelvin
fahrenheit_to_kelvin(32.0)
```

## Nesting functions
```{r}
celsius_to_kelvin(
  fahrenheit_to_kelvin(32.0)
  )
```

C:/Users/user/Google Drive/RWorkshop/Session-I Data Types.Rmd

### Reference:
* https://swcarpentry.github.io/r-novice-inflammation/13-supp-data-structures/

